Carbon Management Strategies for a Climate-Neutral European Economy

### 2023-03-07 report/sections/introduction.tex
update paper

\begin{itemize}
    \item CCUS or CCTUS seem to be covered mainly by IAM community, not energy system modelling. 
    \item \cite{CaptureMapGetSitespecific} Recent Swiss study on CCTS
    \item \cite{weiProposedGlobalLayout2021} Chinese Nature study on global CCS potential: 92 GtCO$_2$ mitigated globally of which 59 Gt will be sequestered. All at cost of 0.12\% of global GDP   
    \item \cite{ToolsGreenTransition} Endravas Capture Map estimates 1.4 Gt capturing potential from point sources in Europe.
    \item SimCCS Optimization software for MIP problems with infrastructure decision \cite{middletonSimCCSOpensourceTool2020}
\end{itemize}

\textbf{Companies and projects}
\begin{itemize}
    \item Potential Carbon Management hubs are evolving Tree Energy, Carbfix, Equinor    
    \item Mission Innovation Hydrogen Valley Platform \cite{H2ValleysMissionInnovation} collects hydrogen flagship projects to facilitate 
    \item European Hydrogen Backbone \cite{gasforclimateEuropeanHydrogenBackbone2022} 
\end{itemize}

\textbf{Legislation}
\begin{itemize}
    \item German climate protection act \cite{KlimaschutzgesetzKlimaneutralitaetBis} targets climate neutrality by 2045, and 65\% CO$_2$ reduction by 2030. (762 Mt CO2-eq emitted in 2021; 438 Mt targeted for 2030) 
    \item Fit-for-55 package by EU targets 55\% reduction by 2030 in the EU
    \item REpowerEU \cite{REPowerEU} targets accelerated energy transition and security. A fast ramp up of a green Hydrogen system should support transition of hard-to-abate sectors. Domestic production should be 10 MT H$_2$ by 2030, same as import. Additional funding for Clean Hydrogen Partnership through Horizon Europe Programme. 
    \item Delegated Act for green H$_2$
    \item 
\end{itemize}

@@

>>>>

\begin{itemize}
    \item CCUS or CCTUS seem to be covered mainly by IAM community, not energy system modelling.
    \item ~\cite{CaptureMapGetSitespecific} Recent Swiss study on CCTS
    \item ~\cite{weiProposedGlobalLayout2021} Chinese Nature study on global CCS potential: 92 GtCO$_2$ mitigated globally of which 59 Gt will be sequestered. All at cost of 0.12\% of global GDP
    \item ~\cite{ToolsGreenTransition} Endravas Capture Map estimates 1.4 Gt capturing potential from point sources in Europe.
    \item SimCCS Optimization software for MIP problems with infrastructure decision ~\cite{middletonSimCCSOpensourceTool2020}
\end{itemize}

\textbf{Companies and projects}
\begin{itemize}
    \item Potential Carbon Management hubs are evolving Tree Energy, Carbfix, Equinor
    \item Mission Innovation Hydrogen Valley Platform ~\cite{H2ValleysMissionInnovation} collects hydrogen flagship projects to facilitate
    \item European Hydrogen Backbone ~\cite{gasforclimateEuropeanHydrogenBackbone2022}
\end{itemize}

\textbf{Legislation}
\begin{itemize}
    \item German climate protection act ~\cite{KlimaschutzgesetzKlimaneutralitaetBis} targets climate neutrality by 2045, and 65\% CO$_2$ reduction by 2030. (762 Mt CO2-eq emitted in 2021; 438 Mt targeted for 2030)
    \item Fit-for-55 package by EU targets 55\% reduction by 2030 in the EU
    \item REpowerEU ~\cite{REPowerEU} targets accelerated energy transition and security. A fast ramp up of a green Hydrogen system should support transition of hard-to-abate sectors. Domestic production should be 10 MT H$_2$ by 2030, same as import. Additional funding for Clean Hydrogen Partnership through Horizon Europe Programme.
    \item Delegated Act for green H$_2$
    \item
\end{itemize}

@@

In net-zero economies, carbon capture is considered a key technology for balancing emissions from hard-to-abate sectors and inevitable emissions from land use and agriculture. Capturing carbon at point sources and from the atmosphere by direct air capture (DAC) opens up new opportunities for producing drop-in fuels such as methane, methanol, or carbon-neutral liquid hydrocarbons, known as Carbon Capture and Utilization (CCU), and for storing carbon in geological formations, known as Carbon Capture and Storage (CCS). Together with carbon transport, these carbon management technologies are considered to be part of what is called the New Carbon Economy~\cite{arniehellerNewCarbonEconomy2019}.

Individual potentials of carbon management technologies have been discussed in detail in the literature and included in previous modeling efforts~\cite{burandtDecarbonizingChinaEnergy2019,caprosEnergysystemModellingEU2019,damoreOptimalDesignEuropean2021,larsonNetZeroAmericaPotential2021,maModelingOptimizationCombined2021,mikulcicFlexibleCarbonCapture2019,williamsCarbonNeutralPathwaysUnited2021}. Recently, policymakers and industry have been committing to carbon management strategies, planning the first infrastructure components, and developing business models for emerging sectors of the economy~\cite{adomaitisEquinorRWEBuild2023,apnewswireGermanyDrawLegislation2023,KohlenstoffKannKlimaschutz2023,OGETESJoin2022,TESHydrogenLife2023}.

However, there is still a lack of careful assessment of the benefits of carbon infrastructure in geographic terms. As we show in our study, the dynamics of carbon capture, transport, use, and storage in different regions are highly dependent on the availability of carbon pipelines and sequestration sinks. To our knowledge, there has never been a study that has examined carbon management with a detailed geographic representation for the future European energy system.

In this paper, we therefore present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-EUR-Sec energy system model and encompasses all relevant sectors. We enable the system to optimize the design of renewable energy sources and storage technologies, as well as the transmission of electricity, hydrogen, methane, and carbon. Our evaluation focuses on the transport dynamics of carbon and hydrogen through their respective networks on the European continent. We also analyze how an energy system with limited annual sequestration potential prioritizes decarbonization and fuel switching in various sectors, and how the construction of carbon networks varies based on different levels of available carbon sequestration potential.

### 2023-03-07 report/sections/methods.tex
adjust layouts to paper with
update paper

Energy system modelling in PyPSA-EUR-sec~\cite{brownSynergiesSectorCoupling2018}

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=\linewidth]{capacity_map_co2network_1000.pdf}
    \caption{Optimal capacities per sector for a sequestration of 1000 Mt/a.}
    \label{fig:capacity_map_noco2network_1000}
\end{figure*}

@@

>>>>

To model the detailed expansion of the carbon network, we use the open-source PyPSA-Eur-Sec~\cite{PyPSAEurSecSectorCoupledOpen2023} model. This is a fully sector-coupled model with high spatial and temporal resolution and detailed transmission infrastructure for electricity, hydrogen, methane, and carbon networks. The model enables joint optimization of investment and operational decisions for generation, storage, conversion, and transmission infrastructures. For our study, we use an approach with 181~geographic regions for Europe and a temporal resolution of 3~hours. These settings allow us to detail the future carbon network and capture the dynamics of material flow between industrial carbon sources and carbon sinks for both use and storage.

PyPSA-Eur-Sec uses exogenous assumptions for the regional energy demand of the power sector, industry, transport, buildings and agriculture. This includes energy demand for shipping and aviation, and non-energy feedstocks for the chemical industry. To take account of climate neutrality targets, we set a binding cap on zero emissions in 2050. Biomass use in each region is limited by its biomass potential, but biomass transport between regions is allowed and the model accounts for transport costs. Furthermore, we limit the expansion of the electric transmission system to 50\% of today's total capacity to account for the difficulties in placing new transmission projects. The geographical layout of process emissions from industry is derived from~\cite{piamanzGeoreferencedIndustrialSites2018}. For carbon sequestration, only offshore sites are considered as potential sinks (see Figure~\ref{fig:carbonSequestrationPotentials}). This is partly because offshore storage potential tends to be greater than onshore potential, and partly because carbon storage infrastructure projects near population centers may be difficult to implement due to public safety concerns. We make conservative assumptions in estimating storage potential by truncating the total potential to 25~Mt per site and calculate annual storage availability when filled over 25 years.
On the carbon utilization side, we include three drop-in fuel production technologies in our model: Methanation, Methanolization, and Fischer-Tropsch (FT) synthesis. For the fuels to be considered carbon-neutral substitutes, they must be produced using carbon from a technology with negative emissions. Synthetic methane can be used as a substitute for natural gas or biogas and is used in combined heat and power (CHP) plants, in gas boilers for home heating, or to meet the gas needs of industry. Synthetic methanol is used in the model to decarbonize fuel demand in the marine industry. Finally, FT fuels can replace fossil oil in the production of naphtha for industry and kerosene for aviation, or be used as machinery oil for agriculture.
All technology cost assumptions for 2050 are listed taken from~\cite{lisazeyenPyPSATechnologydataTechnology2023}.

To explore the impact of carbon sequestration on the optimal layout of carbon management technologies, we vary the global carbon sequestration potential from 200~Mt to 1000~Mt per year in steps of 200~Mt.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{capacity_map_electricity_co2network_1000.pdf}
    \caption{Optimal capacities per of the electricity sector for a sequestration of 1000 Mt/a.}
    \label{fig:capacity_map_noco2network_1000}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{capacity_map_carbon_co2network_1000.pdf}
    \caption{Optimal capacities per of the carbon sector for a sequestration of 1000 Mt/a.}
    \label{fig:capacity_map_noco2network_1000}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{operation_bar_co2network_200.pdf}
\end{figure}
